---
title: "Create_ID_key"
author: "Shelly Trigg"
date: "2/24/2020"
output: html_document
---

read in data
```{r}
KNM_metab <- read.csv("data/mx338976_pteropod_genmetab0917_phxDO.csv", stringsAsFactors = FALSE)

KNM_lipid <- read.csv("data/mx339045_lipid_sampleTRTlipidsonly.csv", stringsAsFactors = FALSE)
```

transpose column names 
```{r}
KNM_metab_colnames <- data.frame(matrix(0,ncol(KNM_metab[,-c(1:2)]),1))
KNM_metab_colnames[,1] <- colnames(KNM_metab[,-c(1:2)])
colnames(KNM_metab_colnames) <- "BinBasename_R"

KNM_lipid_colnames <- data.frame(matrix(0,ncol(KNM_lipid[,-c(1:2)]),1))
KNM_lipid_colnames[,1] <- colnames(KNM_lipid[,-c(1:2)])
colnames(KNM_lipid_colnames) <- "Identifier_R"

```


Read in raw data
```{r}
load libraries
```{r}
library(readxl)
library(ggplot2)
library(gplots)
library(tidyr)
library(dplyr)
library(effsize)
library(broom)



```

read data

```{r}

metab_data <- data.frame(read_xlsx("../abundance_visualization/data/mx 338976 Krista Nichols_pteropods_whole animals_09-2017_submit.xlsx", skip = 8))

lipid_data <- data.frame(read_xlsx("../abundance_visualization/data/mx 339045_Nichols_CSH-QTOF MSMS_lipidomics_09-2017_submit.xlsx", sheet = "submit", skip = 6))

#remove the internal standards
lipid_data <- lipid_data[-grep("iSTD", lipid_data$Annotation),]

#order data by Annotation name
lipid_data <- lipid_data[order(lipid_data$Annotation),]

#for unknown compounds add identifier to annotation column
for (i in 1:nrow(lipid_data)){
  if(is.na(lipid_data$Annotation[i])){
    lipid_data$Annotation[i] <- paste0("X",lipid_data$Identifier[i])
  }
}

```


merge id columns
```{r}
metab_data <- cbind(metab_data[,c("BinBase.name", "BB.id", "PubChem", "KEGG", "InChI.Key")],KNM_metab_colnames)

lipid_data <- cbind(lipid_data[,c("Identifier", "Annotation", "InChI.Key")],KNM_lipid_colnames)
```

Write out ID key table

```{r}
write.table(metab_data,"metab_ID_Key.txt", sep = "\t", row.names = FALSE, quote = FALSE)

write.table(lipid_data,"lipid_ID_Key.txt", sep = "\t", row.names = FALSE, quote = FALSE)

```