---
title: "2016 pteropod survival proportions analysis"
author: "Shelly Trigg"
date: "4/15/2019"
output: rmarkdown::github_document
---

Load libraries
```{r, load_libraries, echo = FALSE}
library(readxl)
library(tidyr)
library(ggplot2)
library(FSA)
library(survminer)
library(survival)
library(lme4)
library(lmerTest)
library(knitr)
```

Read in data
```{r,read_in_data, echo = FALSE}
#survival data
d <- read_xlsx("~/Documents/GitHub/pteropod_pHxDO_metabolomics/survival/pteropod_pHxDO2016_masterdatasheet.xlsx", sheet = "sample IDs living", skip = 2)

#treatment data
treatments <- read.csv("~/Documents/GitHub/Seawater-Chemistry-Analysis/2016-17_PteropodExp_WaterChem/PteropodWaterChem/Treatments.csv", stringsAsFactors = FALSE)
treatments$MOATS <- paste0("M",treatments$MOATS)
```

Format survival data
```{r, format_survival_data, echo = FALSE}
#fix column names which are dates
weird_date <- as.numeric(colnames(d)[3:11], quote = FALSE)
good_dates <- as.Date(weird_date, origin = '1899-12-30')

#convert d to data frame
d <- data.frame(d)

#convert date column names to reformated date strings
colnames(d)[3:11] <- as.character(good_dates)

#rename the first two columns
colnames(d)[1:2] <- c("MOATS", "Jar")

#remove rows containing repetitive date info
d <- d[-grep("^42",d$`2016-11-22`),]

#remove rows containing only NAs
d <- d[which(!is.na(d$`2016-11-22`)),]

#add column for first day everything was put on MOATS
d$`2016-11-21` <- "L1"

#reorder columns so dates are all next to each other
d <- d[,c(1,2,16,3:15)]

#remove MOATS 2 and 9 which had problems
d <- d[-grep("M9", d$MOATS),]


#crab_id column
d$crabID <- paste(d$MOATS, d$Jar, sep = "_")

#reformat date column
STACKED_d <- tidyr::gather(d, date, status, 3:12)

#make event column

for(i in 1:length(STACKED_d$status)){
  if(STACKED_d$status[i] == "L1"){
    STACKED_d$event[i] <- "start"
  }
  if(STACKED_d$status[i] == "D"){
    STACKED_d$event[i] <- "Dead"
  }
  if(STACKED_d$status[i]== "2L"){
    STACKED_d$event[i] <- "unknown"
  }
  if(STACKED_d$status[i] == "M"){
    STACKED_d$event[i] <- "unknown"
  }
  if(STACKED_d$status[i] == "LL"){
    STACKED_d$event[i] <- "unknown"
  }
 if(STACKED_d$status[i] == "L"){
    STACKED_d$event[i] <- "NA"
  }
  if(STACKED_d$status[i] == "----"){
    STACKED_d$event[i] <- "NA"
  }
}
```


remove samples with 2L, M, or LL (2L = 2 animals in one jar (misplacement), M = missing, LL = live but lost..not sure what that actually means)
```{r, echo=FALSE}
bad_ptero_list <- unique(STACKED_d[grep("unknown", STACKED_d$event), "crabID"])

STACKED_d <- STACKED_d[which(!(STACKED_d$crabID %in% bad_ptero_list)),]

#remove columns with no event ("--")

STACKED_d <- STACKED_d[-grep("----", STACKED_d$status),]
```

Add Duration column
```{r, echo = FALSE}
#assign durations
STACKED_d <- STACKED_d[order(STACKED_d$crabID),]
durCounter <- 0
for(i in 1:length(STACKED_d$event)){
  if(STACKED_d$event[i] == "start"){
    durCounter <- 0
  }
  STACKED_d$duration[i] <- durCounter
  durCounter <- durCounter + 1
}

```

Add survival column
```{r, echo = FALSE}
STACKED_d$isDead <- NA
lastDayObs <- "2016-11-30"

for(i in 1:length(STACKED_d$event)){
  if(STACKED_d$event[i] == "Dead"){
    STACKED_d$isDead[i]  <- 1
  }
  if(STACKED_d$status[i] == "L" && STACKED_d$date[i] == lastDayObs){
    STACKED_d$isDead[i] <- 0
    STACKED_d$event[i] <- "AliveAtEnd"
  }

}


```

merge treatment data
```{r, echo = FALSE}

STACKED_d <- merge(STACKED_d, treatments, by = "MOATS")
```

**survival analysis**
```{r, echo = FALSE}
dSub <- STACKED_d[!is.na(STACKED_d$isDead),]
surv <- Surv(time = dSub$duration, event = dSub$isDead, type = "right")
ggsurvplot(survfit(surv ~ Target_Treatment, dSub), risk.table = FALSE, pval = FALSE, conf.int = FALSE,font.main = 16, font.x =  16, font.y = 16, font.tickslab = 16, font.legend = 10, break.time.by = 1,legend = c(0.4, 0.4), legend.title = "Treatment", title = "pHxDO effect on survival", xlab = "Time (Days)", xlim = c(0,9), ylim = c(0.3,1)) 

```


**treatment effect on duration to death**
```{r}
durTreat <- aov(duration ~ Treatment_abbv,data = dSub[which(dSub$isDead == 1),])
summary(durTreat)
```

**MOATS effect on duration to death**
```{r}
durMOATS <- aov(duration ~ MOATS,data = dSub[which(dSub$isDead == 1),])
summary(durMOATS)
```

**Test significance of treatment on duration to death with MOATS as a random effect**
```{r}
fitME <- lmer(duration~ Treatment_abbv + (1|MOATS), data = dSub[which(dSub$isDead == 1),])
summary(fitME)
```


**plot survival proportions for treatments**
```{r, echo = FALSE}
ggplot(dSub) + geom_bar(aes(Treatment_abbv, fill = event))
```

**plot survival proportions (fraction) for treatments**
```{r, echo = FALSE}
ggplot(dSub) + geom_bar(aes(Treatment_abbv, fill = event), position = "fill")
```


**plot survival proportions for MOATS**
```{r, echo = FALSE}
ggplot(dSub) + geom_bar(aes(MOATS, fill = event), position = "fill") + facet_wrap(~Treatment_abbv)
```

**calculate chi square and p values for proportions**
```{r, echo = FALSE}
comparisons <- c("HH_HL", "HH_LH", "HH_LL", "HL_LH", "HL_LL", "LH_LL")
surv_stats <- data.frame()
for(i in 1:length(comparisons)){
  x <- c(length(dSub[which(dSub$isDead == 0 & dSub$Treatment_abbv == substr(comparisons[i], 1,2)),"isDead"]), length(dSub[which(dSub$isDead == 0 & dSub$Treatment_abbv == substr(comparisons[i], 4,5)),"isDead"]))
  n <- c(length(dSub[which(dSub$Treatment_abbv == substr(comparisons[i], 1,2)),"isDead"]), length(dSub[which(dSub$Treatment_abbv == substr(comparisons[i], 4,5)),"isDead"]))
  chi <- prop.test(x = x, n = n, correct = FALSE)
  row <- data.frame(t(data.frame(c(comparisons[i], chi$statistic, chi$p.value))))
  colnames(row) <- c("Comparison", "ChiSq", "P.value")
  surv_stats <- rbind(surv_stats, row)
}
rownames(surv_stats) = NULL

surv_stats$P.value_bonferroni <- p.adjust(surv_stats$P.value, method = "bonferroni")
```

```{r}
kable(surv_stats, caption = "Treatment effect on survival proportions test table")
```

## EXCLUDING MOATS 2

**survival analysis**
```{r, surv_analysis_noM2, echo = FALSE}
dSub <- STACKED_d[which(!is.na(STACKED_d$isDead) & STACKED_d$MOATS != "M2"),]
surv <- Surv(time = dSub$duration, event = dSub$isDead, type = "right")
ggsurvplot(survfit(surv ~ Target_Treatment, dSub), risk.table = FALSE, pval = FALSE, conf.int = FALSE,font.main = 16, font.x =  16, font.y = 16, font.tickslab = 16, font.legend = 10, break.time.by = 1,legend = c(0.4, 0.4), legend.title = "Treatment", title = "pHxDO effect on survival", xlab = "Time (Days)", xlim = c(0,9), ylim = c(0.3,1)) 
```


**treatment effect on duration to death (without M2)**
```{r}
durTreat <- aov(duration ~ Treatment_abbv,data = dSub[which(dSub$isDead == 1),])
summary(durTreat)
```

**MOATS effect on duration to death (without M2)**
```{r}
durMOATS <- aov(duration ~ MOATS,data = dSub[which(dSub$isDead == 1),])
summary(durMOATS)
```

**Test significance of treatment on duration to death with MOATS as a random effect (without M2)**
```{r}
fitME <- lmer(duration~ Treatment_abbv + (1|MOATS), data = dSub[which(dSub$isDead == 1),])
summary(fitME)
```


**plot survival proportions for treatments (without M2)**
```{r, echo =FALSE}
ggplot(dSub) + geom_bar(aes(Treatment_abbv, fill = event))
```

**plot survival proportions (as fractions) for treatments (without M2)**
```{r, echo = FALSE}
ggplot(dSub) + geom_bar(aes(Treatment_abbv, fill = event), position = "fill")
```


**plot survival proportions for MOATS (without M2)**
```{r, echo = FALSE}
ggplot(dSub) + geom_bar(aes(MOATS, fill = event), position = "fill") + facet_wrap(~Treatment_abbv)
```

**calculate chi square and p values for proportions (without M2)**
```{r, echo = FALSE}
comparisons <- c("HH_HL", "HH_LH", "HH_LL", "HL_LH", "HL_LL", "LH_LL")
surv_stats <- data.frame()
for(i in 1:length(comparisons)){
  x <- c(length(dSub[which(dSub$isDead == 0 & dSub$Treatment_abbv == substr(comparisons[i], 1,2)),"isDead"]), length(dSub[which(dSub$isDead == 0 & dSub$Treatment_abbv == substr(comparisons[i], 4,5)),"isDead"]))
  n <- c(length(dSub[which(dSub$Treatment_abbv == substr(comparisons[i], 1,2)),"isDead"]), length(dSub[which(dSub$Treatment_abbv == substr(comparisons[i], 4,5)),"isDead"]))
  chi <- prop.test(x = x, n = n, correct = FALSE)
  row <- data.frame(t(data.frame(c(comparisons[i], chi$statistic, chi$p.value))))
  colnames(row) <- c("Comparison", "ChiSq", "P.value")
  surv_stats <- rbind(surv_stats, row)
}
rownames(surv_stats) = NULL

surv_stats$P.value_bonferroni <- p.adjust(surv_stats$P.value, method = "bonferroni")
```

```{r}
kable(surv_stats, caption = "Treatment effect on survival proportions test table (without M2)")
```
